A comparative analysis of penalized linear mixed models in structured genetic data

### 2021-02-14 main.tex
Making DAGs black and white
Sec 5. Responding to: penalty parameter selection; also a tough one. We need to say more, but not a great deal is known.

One method for analyzing data in the presence of correlated loci is LD pruning, which involves sequentially thinning SNPs that exceed a correlation threshold, so that only variants that are relatively independent are used in modeling and testing procedures.
%LD pruning can be an important preprocessing step in a GWAS analysis, for which there exist a variety of algorithms and methods.
An alternative solution is the use of multivariate methods such as multiple linear regression, which adjusts for and estimates the effects of numerous variants simultaneously. However, classical multivariate approaches rely on data with more observations than unknown parameters. This is typically infeasible in genome-wide studies where the number of variants greatly exceeds the number of individuals. Penalized regression methods make such problems tractable by constraining the solution space of coefficient estimates. The lasso is a widely used penalized regression method \citep{tibshirani1996regression}. It relies on an assumption of sparsity; in other words, that many features will have no effect. Thus, although the total number of features may be large, the number that actually affect the outcome will be small. This allows coefficient estimation and variable selection to be accomplished simultaneously, and provides considerable computational advantages that scale up well to very large data sets. These features make the lasso attractive for the analysis of high-dimensional genetic data where identifying the variants most highly associated with a particular trait is often of interest. Although separately testing each SNP is still the most common way of analyzing GWAS data, the use of lasso and related methods is increasing in use due to computational advances that make it practical to apply on a genome-wide scale \citep{qian2020fast, prive2018efficient}. \anna{More background on lasso}

@@

\begin{equation}
\centering
\begin{tikzpicture}[baseline=(current  bounding  box.center)]
    \node (1) at (0,0) {Population Stratification};
    \node (2) [right = of 1] {Environment};
    \node (3) [below = of 1] {Observed SNPs};
    \node (4) [below = of 2] {Phenotype};
    \path[bidirected, -] (1) edge (2);
    \path[red, line width = 0.5] (1) edge (3);
    \path (2) edge (4);
    \path (3) edge (4);
    \tkzMarkSegment[color=red,pos=.5,mark=x](1,3)
\end{tikzpicture}
%\caption{The causal pathway between the environment and observed SNPs may be indirectly blocked by many methods that attempt to reduce confounding due to population structure.}
\label{diag:ps_env_block}
\end{equation}

@@

\begin{equation}
\centering
\begin{tikzpicture}[baseline=(current  bounding  box.center)]
    \node (1) at (0,0) {Population Stratification};
    \node (2) [right = of 1] {Environment};
    \node (3) [below = of 1] {Observed SNPs};
    \node (4) [below = of 2] {Phenotype};
    \path[bidirected, -, red] (1) edge (2);
    \path (1) edge (3);
    \path (2) edge (4);
    \path (3) edge (4);
    \tkzMarkSegment[color=red,pos=.5,mark=x](1,2)
\end{tikzpicture}
%\caption{The causal pathway between the environment and observed SNPs may be blocked if genetic information does not serve as an accurate proxy for it.}
\label{diag:ps_env_block2}
\end{equation}

@@

%------------------------------%
\subsection{Software}
\label{sec:software}
\anna{More discussion of software. Add a small section? R-based Software for lasso, pc-lasso, ggmix, plmm, PLINK functions, KING, snpRelate}

@@

\anna{penalty parameter selection; also a tough one. We need to say more, but not a great deal is known.}

Although cross-validation and information criterion-based methods are commonly used to select $\lambda$ values, in this context they have the potential to result in misleading comparisons. Because lasso, PC-lasso, and LMM-lasso use different likelihoods, an information-criterion-based approach such as BIC would not lend itself to comparable results across methods. Meanwhile, whether cross-validation is advisable or not in the presence of dependent samples is an open question \citep{roberts2017cross}.

A total of 1000 simulations were performed for each unique combination of parameters, datatype, subpopulation structure, and environmental effect structure.

@@

>>>>

One method for analyzing data in the presence of correlated loci is LD pruning, which involves sequentially thinning SNPs that exceed a correlation threshold, so that only variants that are relatively independent are used in modeling and testing procedures.
%LD pruning can be an important preprocessing step in a GWAS analysis, for which there exist a variety of algorithms and methods.
An alternative solution is the use of multivariate methods such as multiple linear regression, which adjusts for and estimates the effects of numerous variants simultaneously. However, classical multivariate approaches rely on data with more observations than unknown parameters. This is typically infeasible in genome-wide studies where the number of variants greatly exceeds the number of individuals. Penalized regression methods make such problems tractable by constraining the solution space of coefficient estimates. The lasso is a widely used penalized regression method \citep{tibshirani1996regression}. \anna{More background on lasso} It relies on an assumption of sparsity; in other words, that many features will have no effect. Thus, although the total number of features may be large, the number that actually affect the outcome will be small. This allows coefficient estimation and variable selection to be accomplished simultaneously, and provides considerable computational advantages that scale up well to very large data sets. These features make the lasso attractive for the analysis of high-dimensional genetic data where identifying the variants most highly associated with a particular trait is often of interest. Although separately testing each SNP is still the most common way of analyzing GWAS data, the use of lasso and related methods is increasing in use due to computational advances that make it practical to apply on a genome-wide scale \citep{qian2020fast, prive2018efficient}.

@@

\begin{equation}
\centering
\begin{tikzpicture}[baseline=(current  bounding  box.center)]
    \node (1) at (0,0) {Population Stratification};
    \node (2) [right = of 1] {Environment};
    \node (3) [below = of 1] {Observed SNPs};
    \node (4) [below = of 2] {Phenotype};
    \path[bidirected, -] (1) edge (2);
    \path[line width = 0.5] (1) edge (3);
    \path (2) edge (4);
    \path (3) edge (4);
    \tkzMarkSegment[pos=.5,mark=x](1,3)
\end{tikzpicture}
%\caption{The causal pathway between the environment and observed SNPs may be indirectly blocked by many methods that attempt to reduce confounding due to population structure.}
\label{diag:ps_env_block}
\end{equation}

@@

\begin{equation}
\centering
\begin{tikzpicture}[baseline=(current  bounding  box.center)]
    \node (1) at (0,0) {Population Stratification};
    \node (2) [right = of 1] {Environment};
    \node (3) [below = of 1] {Observed SNPs};
    \node (4) [below = of 2] {Phenotype};
    \path[bidirected, -] (1) edge (2);
    \path (1) edge (3);
    \path (2) edge (4);
    \path (3) edge (4);
    \tkzMarkSegment[pos=.5,mark=x](1,2)
\end{tikzpicture}
%\caption{The causal pathway between the environment and observed SNPs may be blocked if genetic information does not serve as an accurate proxy for it.}
\label{diag:ps_env_block2}
\end{equation}

@@

%------------------------------%
\subsection{Software}
\label{sec:software}
\anna{More discussion of software. Add a small section? R-based Software for lasso, pc-lasso, ggmix, plmm, PLINK functions, KING, snpRelate. Remove other mentions of software in other places and put them here}

@@

Although cross-validation (CV) and information criterion-based methods are commonly used to select $\lambda$ values, in this context they have the potential to result in misleading comparisons. Because lasso, PC-lasso, and LMM-lasso use different likelihoods, an information-criterion-based approach such as BIC would not lend itself to comparable results across methods. Meanwhile, whether cross-validation is advisable or not in the presence of dependent samples is an open question \citep{roberts2017cross}. \anna{We include a more detailed discussion of methods of penalty parameter selection for penalized LMMs to section \ref{sec:discussion}.}

A total of 1000 simulations were performed for each unique combination of parameters, datatype, subpopulation structure, and environmental effect structure. \anna{maybe cut this and include in figure/table captions instead since they requested that for fig1/tab1}

%------------------------------%
\subsection{\anna{Genomewide case study}}
%------------------------------%

@@

\anna{\textbf{penalty parameter selection; also a tough one. We need to say more, but not a great deal is known.}}

The nature of the simulations and analyses in this work have allowed us to select penalty parameters based on the true number of causal features in a model, which allows for fair comparison across methods. In a real data scenario, the true number of causal SNPs would not be known a priori. As such, appropriate methods for penalty parameter selection in penalized LMMs are worth additional discussion. As of yet there is no clear consensus on the optimal way to select a penalty parameter for penalized LMMs.

Several authors have used the Bayesian information criterion (BIC) for variable selection in mixed models \cite{bondell2010joint, schelldorfer2011estimation}. However it has been shown that the Akaike information criterion (AIC) and BIC may perform poorly in the high-dimensional setting \citep{fan2013tuning}. The high-dimensional BIC introduced by \citet{fan2013tuning} penalizes model parameters more severely in an attempt to perform better in high dimensional settings. \citep{bhatnagar2020simultaneous} employs this high-dimensional BIC as its default method to select $\lambda$.

Subsampling approaches, such as CV and stability selection, are another common class of variable selection methods often preferred to information criterion-based approaches. \anna{cite? It seems like you see CV done way more often...} \citet{Rakitsch2012} used stability selection to evaluate the significance of individual SNPs \citep{meinshausen2010stability}. \citet{guo2019combining} used a combination of both methods for feature selection, first selecting several $\lambda$ values via CV, and then using these penalty parameters for stability selection and final feature selection.

As previously mentioned, however, the advisability of subsampling methods in the presence of correlated observations remains an open question \citep{roberts2017cross, jia2012preconditioning}. A rigorous assessment of penalty parameter selection in the context of penalized LMMs is beyond the scope of this work, but worthy of further investigation.

\anna{I don't think we should bring up the idea of CV/subsampling on original vs transformed data in this paper at all - I think it's just confusing without a lot of additional explanation.}
